\section{Background}
In contrast to life, an alignment can end in three different ways:
\begin{enumerate}
  \I \begin{tabular}{c}$\mathtt{n}$\\$\mathtt{-}$\end{tabular}
  \I \begin{tabular}{c}$\mathtt{n}$\\$\mathtt{n}$\end{tabular}
  \I \begin{tabular}{c}$\mathtt{-}$\\$\mathtt{n}$\end{tabular}
\end{enumerate}

where $\tt n$ indicates any residue. These three possibilities suggest
the following recursion for the number of possible global alignments
\cite{wat95:int}:
\[
f(n,k) = f(n-1,k)+f(n-1,k-1)+f(n,k-1),
\]
where $f(n,k)$ is the number of possible alignments between two
strings of length $n$ and $k$. If we further define the boundary
condition that there is a single alignment between a sequence of any
length and a sequence of length zero, we are in a position to compute
the number of possible global alignments by dynamic programming.

The possible number of local alignments is the sum of the number of global
alignments for all possible pairs of subsequences. A sequence of
length $n$ contains $n-m+1$ subsequence of length $m$. Consider as an
example two sequences of length two each. Nine pairs of subsequences
can be formed between these. Four of the subsequence pairs have length (1,1), yielding three possible global
alignments each; another four have length (2,1) or (1,2), yielding
five possible global alignments each; one has length (2,2), yielding
13 possible global alignments. The number of possible local alignments
is therefore $12 + 20 + 13 = 45$.

\section{Description of Program}
The program uses dynamic programming to calculate the number of
different global or local alignments that
can be constructed for two sequences of lengths between 1 and 400. The
user can choose the lengths of the two sequences as well as the
alignment mode (global or local).

\section{Tutorial}
\begin{itemize}
  \I Contemplate the number of global alignments that can be
  formed between two sequences of lengths 339 and 320.
  \I Compare that number to the number of molecules contained in a liter
  of water: $3.3 \times 10^{25}$. How much water is needed to
  accumulate the same number of water molecules as there are possible
  alignments between two sequences of lengths 339 and 320?
  \I Compute the time it would take to list all possible global
  alignments between two sequences of length 320 given that the
  calculation of a single alignment takes, say, $10^{-15}$ seconds.
  \I Compute the time it would take to list all possible local
  alignments for two sequences of length 320.
\end{itemize}





